Quantifying the Relationship Between Genetic Diversity and Population Size Suggests Natural Selection Cannot Explain Lewontin's Paradox

### 2021-08-11 manuscript/elife/elife-manuscript.tex
more edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \title{Quantifying Lewontin's Paradox Suggests Natural Selection is Unlikely to Explain the Narrow Range of Diversity Among Species}
\title{Quantifying the Relationship Between Genetic Diversity and Population Size Suggests Natural Selection Cannot Explain Lewontin's Paradox}

@@

Using a phylogenetic mixed-effects model
\citep{Lynch1991-tp,Hadfield2010-ql,De_Villemereuil2014-kt} implemented in Stan
\citep{Carpenter2017-xi,Stan_Development_Team2020-ea}, I estimated the linear
relationship between diversity and population size (on a log-log scale)
accounting for phylogeny, for the 166 taxa without missing data and present in
the synthetic chronogram. This phylogenetic mixed-effects model can account for
correlated residuals among more closely related species when estimating the
relationship between $N_c$ and $\pi$, but does not assume there is phylogenetic
signal in either of these variables. Since the phylogenetic mixed-effects model
simultaneously estimates the degree of phylogenetic structure in the residuals
while fitting the relationship between $N_c$ and $\pi$, this model's estimates
would match those found by ordinary least squares if the residuals were
distributed independently. Overall, this approach is conservative, making no
assumptions about the source of the phylogenetic signal while accounting for
violations of the regression model due to dependence among the residuals (see
\cite{Revell2010-mf} for a discussion of this).

@@

Since the phylogenetic mixed-effects model simultaneously estimates the
variance of the phylogenetic effect ($\sigma_p^2$) and the residual variance
($\sigma_r^2$), these can be used to estimate the phylogenetic signal, $\lambda
= \nicefrac{\sigma_p^2}{(\sigma_p^2 + \sigma_r^2)}$
(\cite{Lynch1991-tp,De_Villemereuil2014-kt}; see \cite{Freckleton2002-ly} for a
comparison to Pagel's $\lambda$). If the residuals were free of correlations
due to shared phylogenetic history, then $\lambda = 0$ and all the variance
could be explained by evolution on the tips; this is analogous to Lynch's
conjecture that coalescence times should be free of phylogenetic signal
(\citeyear{Lynch2011-qv}). In the relationship between population size and
diversity, the posterior mean of $\lambda = 0.67$ (90\% credible interval
$[0.58, 0.75]$) indicates a majority of the variance perhaps might be
due to shared phylogenetic history (\FIG{pcm}B).

@@

Finally, the poor fit between observed and predicted levels of diversity across
species is not remedied by stronger selection parameters. In
\FIGSUPP[linksel]{linked-sel-alt}B, I increase both selection parameters $U$
and $\gamma$ ten-fold each, and find the same qualitative pattern: on a log-log
scale the relationship between $N_c$ and $\pi$ is linear, while the predicted
diversity under linked selection is non-linear with $N_c$. Under this ten-fold
higher selection regime, there is more overlap between observed and predicted
levels of diversity, but diversity is severely \emph{under-predicted} for
high-$N_c$ species. Additionally, this would imply that selection in
low-to-mid-$N_c$ species is ten-folder higher than estimated in
\emph{Drosophila melanogaster}, which is implausible. Overall, predicted
levels of diversity under models of recurrent hitchhiking and background
selection fit the observed relationship between $N_c$ and $\pi$ poorly, and
this is not overly sensitive to parameter choice.

@@

First, Lynch has correctly pointed out that while coalescence times are much
less than divergence times and should be free of phylogenetic shared history,
the factors that determine coalescence times (e.g.  mutation rates and
effective population size) may not be (2011). In other words, coalescence times
are free from phylogenetic shared history \emph{were we to condition} on these
causal factors that could be affected by shared phylogenetic history. My
estimates of phylogenetic signal in diversity, by contrast, are not conditioned
on these factors. Importantly, even ``correcting for" phylogeny implicitly
favors certain causal interpretations over others
\citep{Westoby1995-wz,Uyeda2018-wf}.  Future work could try to untangle what
causal factors determine coalescence times across species, as well as how these
factors evolve across macroevolutionary timescales. Second, it is a
misconception that a fast rate of trait evolution necessarily reduces
phylogenetic signal \citep{Revell2008-in}, and that if either or both variables
in a regression are free of phylogenetic signal, PCMs are unnecessary
\citep{Revell2010-mf,Uyeda2018-wf}. The evidence of high phylogenetic signal
found in this study suggests PCMs are necessary when fitting the relationship
between $N_c$ and $\pi$ in order to account for correlated residuals among
closely-related species, and to avoid spurious results from phylogenetic
pseudoreplication.

@@

Lewontin's Paradox describes the extent to which the effective population sizes
implied by diversity, $\widetilde{N}_e$, diverge from census population sizes.
However, there are a variety other effective population size estimates
calculable from different data and summary statistics
\citep{Wang2016-mi,Caballero1994-ao,Caballero2020-wm,Galtier2020-fb}. These
include estimators based on the site frequency spectrum, observed decay in
linkage disequilibrium, or temporal estimators that use the variance in allele
frequency change. These alternate estimators capture summaries of the effective
population size on shorter timescales than coalescent-based estimators
\citep{Wang2005-wy}, and thus could be used to tease apart processes that
impact the $N_e$-$N_c$ relationship in the more recent past.

Temporal $N_e$ estimators already play an important role in understanding
another summary of the $N_e$-$N_c$ relationship: the ratio
$\nicefrac{N_e}{N_c}$, which is an important quantity in conservation genetics
\citep{Frankham1995-xz,Mace1991-ai} and in understanding evolution in
highly fecund marine species. Surveys of the short-term $\nicefrac{N_e}{N_c}$
relationship across taxa indicate mean $\nicefrac{N_e}{N_c}$ is on order of
$\approx 0.1$ \citep{Frankham1995-xz,Palstra2008-rt,Palstra2012-vy}, though
the uncertainty in these estimates is high, and some species with sweepstakes
reproduction systems like Pacific Oyster (\emph{Crassostrea gigas}) can have
$\nicefrac{N_e}{N_c} \approx 10^{-6}$. Estimates of the $\nicefrac{N_e}{N_c}$
ratio are an important, yet under appreciated piece of solving Lewontin's
Paradox. For example, if $N_e$ is estimated from the allele frequency change
across a single generation (i.e. \cite{Waples1989-sj}), $\nicefrac{N_e}{N_c}$
constrains the variance in reproductive success
\citep{Wright1938-tv,Nunney1993-ef,Nunney1996-wy}. This implies that apart
from species with sweepstakes reproductive systems, the variance in
reproductives success each generation (whether heritable or non-heritable) is
likely insufficient to significantly contribute to constraining
$\widetilde{N}_e$ for most taxa. Still, further work is needed to characterize
(1) how $\nicefrac{N_e}{N_c}$ varies with $N_c$ across taxa (though see
\cite{Palstra2012-vy}, Figure 2), and (2) the variance of $\nicefrac{N_e}{N_c}$
over longer time spans (i.e. how periodic sweepstakes reproductive events act
to constrain $N_e$). Overall, characterizing how $\nicefrac{N_e}{N_c}$ varies
across taxa and correlates with ecology and life history traits could provide
clues into the mechanisms that leads propagule size and survivorship curves to
be predictive of diversity levels across taxa
\citep{Romiguier2014-bp,Hallatschek2018-iu,Barry2020-pp}.

Finally, short-term temporal $N_e$ estimators may play an important role in
resolving Lewontin's Paradox. These estimators, along with short-term estimates
of the impact of linked selection \citep{Buffalo2019-qs,Buffalo2020-my},
can inform us how much diversity is depressed across shorter timescales, free
from the rare strong selective events or severe bottlenecks that impact
pairwise diversity. It could be that in any one generation, selection
contributes more to the variance of allele frequency changes than drift, yet
across-taxa patterns in diversity are better explained processes acting
sporadically on longer timescales, such as colonization, founder effects, and
bottlenecks. Thus, the pairwise diversity may not give us the best picture of
the generation to generation evolutionary processes acting in a population to
change allele frequencies. Furthermore, certain observed adaptations are
inexplicable given implied long-term coalescent effective population sizes, and
are only possible if short-term effective population sizes are orders of
magnitude larger \citep{Karasov2010-fb,Barton2010-pi}.

@@

In \emph{Building a Science of Population Biology}
(\citeyear{Lewontin2004-fb}), Lewontin laments the difficulty of uniting
population genetics and population ecology into a cohesive discipline of
population biology. Lewontin's Paradox of Variation remains a critical unsolved
problem at the nexus of these two different disciplines: across species, we
fail to understand the processes that connect a central parameter of population
ecology, census size, to a central parameter of population genetics, effective
population size. Given that selection seems to fall short in explaining
Lewontin's Paradox, a full resolution will require a mechanistic understanding
the ecological, life history, and macroevolutionary processes that connect
$N_c$ to $N_e$ across taxa. While I have focused exclusively on metazoan taxa
since their population densities are more readily approximated from body mass,
a full resolution must also include plant species (with the added difficulties
of variation in selfing rates, different dispersal strategies, pollination,
etc.).

Looking at Lewontin's Paradox through an macroecological and macroevolutionary
lens begets interesting questions outside of the traditional realm of
population genetics.  Here, I have found that diversity and $N_c$ have a
surprisingly consistent relationship without many outliers, despite the wildly
disparate ecologies, life histories, and evolutionary histories of the taxa
included. Furthermore, taxa with very large census sizes have surprisingly low
diversity. Is this explained by macroevolutionary processes, such as different
rates of speciation for large-$N_c$ taxa? Or, are the levels of diversity we
observe today an artifact of our timing relative to the last glacial maximum,
or the last major extinction? Did large-$N_c$ prehistoric animal populations
living in other geological eras have higher levels of diversity than our
present taxa? Or, does ecological competition occur on shorter timescales such
that strong population size contractions transpire and depress diversity, even
if a species is undisturbed by climatic shifts or mass extinctions? Overall,
patterns of diversity across taxa are determined by many overlaid evolutionary
and ecological processes occurring on vastly different timescales. Lewontin's
Paradox of Variation may persist unresolved for some time because the
explanation requires synthesis and model building at the intersection of all
these disciplines.

@@

\begin{align}
  \E(\pi) &= \frac{\theta}{\theta + \nicefrac{1}{B} + 2NS} \\
          &\approx \frac{\theta}{\nicefrac{1}{B} + 2NS} 
\end{align}
%
(cf. equation 1 \cite{Elyashiv2016-vt}, and equation 20 of \cite{Coop2012-cd}).
The BGS component is given by \cite{Hudson1995-xc},

@@

\begin{align}
  S = \frac{\nu_\text{BP}}{r_\text{BP}} J
\end{align}
%
(cf. \cite{Coop2012-cd} equation 20) where $J$ is the probability that two
lineages coalesce down to one, given sweeps occur uniformly along the genome.
Under this homogeneous sweep model, $J$ is

@@

\begin{align}
  S(x) = \frac{1}{T} \sum_{i_S} \alpha(i_S) \sum_{y \in a(i_S)} \int \exp(-r(x, y) \tau(s, N)) g(s | i_S) ds
\end{align}
%
where $T$ is the number of generations that substitutions accrue, $i_S = 1,
\ldots, I_S$ is the annotation class (e.g. exons, introns, UTRs), $\alpha(i_S)$
is the fraction of substitutions in annotation class $i_S$ that are beneficial,
$a(i_S)$ is the set of all substitutions in annotation class $i_S$, $\tau(s,
N)$ is the fixation time of a site with additive effect $s$, and $g(s | i_S)$
is the distribution of selection coefficients for annotation class $i_S$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \title{Quantifying Lewontin's Paradox Suggests Natural Selection is Unlikely to Explain the Narrow Range of Diversity Among Species}

\title{Quantifying the Relationship Between Genetic Diversity and Population Size Suggests Natural Selection Cannot Explain Lewontin's Paradox}

@@

Using a phylogenetic mixed-effects model
\citep{Lynch1991-tp,Hadfield2010-ql,De_Villemereuil2014-kt} implemented in Stan
\citep{Carpenter2017-xi,Stan_Development_Team2020-ea}, I estimated the linear
relationship between diversity and population size (on a log-log scale)
accounting for phylogeny, for the 166 taxa without missing data and present in
the synthetic chronogram. This type of model is needed because closely-related
species may differ from the average trend between $N_c$ and $\pi$ in similar
ways due to shared phylogenetic history, similar life history traits, etc., and
thus do not represent independent observations as is assumed by the standard
regression model. This is a form of phylogenetic pseudoreplication, and can be
accounted for with a phylogenetic mixed-effects model. The phylogenetic
mixed-effects model does not assume that there is phylogenetic structure in
either $N_c$ or $\pi$ (which itself is not a violation of the standard
regression model, \cite{Revell2010-mf} and \cite{Uyeda2018-wf}), but rather
accounts for phylogenetic correlations structure in the residuals if any is
present. Importantly, phylogenetic mixed-effects models simultaneously estimate
the degree of phylogenetic structure in the residuals while fitting the
relationship between $N_c$ and $\pi$. If the residuals are distributed
independently, the estimated relationship would be similar to that found by
ordinary least squares, and the estimated phylogenetic signal would be zero.
Overall, this approach is conservative, making no assumptions about the source
of the phylogenetic signal while accounting for violations of the regression
model due to dependence among the residuals if present (see
\cite{Revell2010-mf} for a discussion of this).

@@

Since the phylogenetic mixed-effects model simultaneously estimates the
variance of the phylogenetic effect ($\sigma_p^2$) and the residual variance
($\sigma_r^2$), these can be used to estimate the phylogenetic signal, $\lambda
= \nicefrac{\sigma_p^2}{(\sigma_p^2 + \sigma_r^2)}$
(\cite{Lynch1991-tp,De_Villemereuil2014-kt}; see \cite{Freckleton2002-ly} for a
comparison to Pagel's $\lambda$). When residuals are free of correlations due
to shared phylogenetic history, then $\lambda = 0$ and all the variance could
be explained by evolution or noise on the tips. In the relationship between
population size and diversity, the posterior mean of $\lambda = 0.67$ (90\%
credible interval $[0.58, 0.75]$) indicates a majority of the variance perhaps
might be due to shared phylogenetic history (\FIG{pcm}B).

@@

Finally, the poor fit between observed and predicted levels of diversity across
species is not remedied by stronger selection parameters. In
\FIGSUPP[linksel]{linked-sel-alt}B, I increase both selection parameters $U$
and $\gamma$ ten-fold each, and find the same qualitative pattern: on a log-log
scale the relationship between $N_c$ and $\pi$ is linear, while the predicted
diversity under linked selection is non-linear with $N_c$. Under this ten-fold
higher selection regime, there is more overlap between observed and predicted
levels of diversity, but diversity is severely \emph{under-predicted} for
high-$N_c$ species. Additionally, this would imply that selection in
low-to-mid-$N_c$ species is ten-folder higher than estimated in
\emph{Drosophila melanogaster}, which is implausible. Overall, this suggests
that present models of linked selection, even with very strong selection across
species, are qualitatively incapable of matching the observed relationship
between $N_c$ and $\pi$ and thus cannot explain Lewontin's Paradox.

@@

First, Lynch has correctly pointed out that while coalescence times are much
less than divergence times and should be free of phylogenetic shared history,
the factors that determine coalescence times (e.g.  mutation rates and
effective population size) may not be (2011). In other words, coalescence times
are free from phylogenetic shared history \emph{were we to condition} on these
causal factors that could be affected by shared phylogenetic history. My
estimates of phylogenetic signal in the residuals, by contrast, are not
conditioned on these factors. Importantly, even ``correcting for" phylogeny
implicitly favors certain causal interpretations over others
\citep{Westoby1995-wz,Uyeda2018-wf}.  Future work could try to untangle what
causal factors determine coalescence times across species, as well as how these
factors evolve across macroevolutionary timescales. Second, it is a
misconception that a fast rate of trait evolution necessarily reduces
phylogenetic signal \citep{Revell2008-in}, and that if either or both variables
in a regression are free of phylogenetic signal, PCMs are unnecessary
\citep{Revell2010-mf,Uyeda2018-wf}. The evidence of high phylogenetic signal
found in this study suggests PCMs are necessary when fitting the relationship
between $N_c$ and $\pi$ in order to account for correlated residuals among
closely-related species, and to avoid spurious results from phylogenetic
pseudoreplication.

@@

Lewontin's Paradox describes the extent to which the effective population sizes
implied by diversity, $\widetilde{N}_e$, diverge from census population sizes.
However, there are a variety other effective population size estimators
calculable from different data and summary statistics
\citep{Wang2016-mi,Caballero1994-ao,Caballero2020-wm,Galtier2020-fb}. These
include estimators based on the site frequency spectrum, observed decay in
linkage disequilibrium, or temporal estimators that use the variance in allele
frequency change through time. These various estimators capture different
summaries of effective population size on shorter timescales than
coalescent-based estimators (see \cite{Wang2005-wy} for a review), and thus
could be used to tease apart processes that impact the $N_e$-$N_c$ relationship
in the more recent past.

Temporal $N_e$ estimators already play an important role in understanding
another summary of the $N_e$-$N_c$ relationship: the ratio
$\nicefrac{N_e}{N_c}$, which is an important quantity in conservation genetics
\citep{Frankham1995-xz,Mace1991-ai} and in understanding evolution in highly
fecund marine species. Surveys of the short-term $\nicefrac{N_e}{N_c}$
relationship across taxa indicate mean $\nicefrac{N_e}{N_c}$ is on order of
$\approx 0.1$ \citep{Frankham1995-xz,Palstra2008-rt,Palstra2012-vy}, though the
uncertainty in these estimates is high, and some species with sweepstakes
reproduction systems like Pacific Oyster (\emph{Crassostrea gigas}) can have
$\nicefrac{N_e}{N_c} \approx 10^{-6}$ \citep{Hedgecock1994-gs}. Estimates of
the $\nicefrac{N_e}{N_c}$ ratio may be an important, yet under appreciated
piece of solving Lewontin's Paradox. For example, if $N_e$ is estimated from
the allele frequency change across a single generation (i.e.
\cite{Waples1989-sj}), $\nicefrac{N_e}{N_c}$ constrains estimates of the
variance in reproductive success
\citep{Wright1938-tv,Nunney1993-ef,Nunney1996-wy}. This implies that apart from
species with sweepstakes reproductive systems, the variance in reproductive
success each generation (whether heritable or non-heritable) is likely
insufficient to significantly contribute to constraining $\widetilde{N}_e$ for
most taxa. Still, further work is needed to characterize (1) how
$\nicefrac{N_e}{N_c}$ varies with $N_c$ across taxa (though see
\cite{Palstra2012-vy}, Figure 2), and (2) the variance of $\nicefrac{N_e}{N_c}$
over longer time spans (i.e. how periodic sweepstakes reproductive events act
to constrain $N_e$). Overall, characterizing how $\nicefrac{N_e}{N_c}$ varies
across taxa and correlates with ecology and life history traits could provide
clues into the mechanisms that leads propagule size and survivorship curves to
be predictive of diversity levels across taxa
\citep{Romiguier2014-bp,Hallatschek2018-iu,Barry2020-pp}.

Finally, short-term temporal $N_e$ estimators may play an important role in
resolving Lewontin's Paradox. These estimators, along with short-term estimates
of the impact of linked selection \citep{Buffalo2019-qs,Buffalo2020-my}, can
inform us how much diversity is depressed by selection on shorter timescales,
free from the rare strong selective events or severe bottlenecks that impact
pairwise diversity. It could be that in any one generation, selection
contributes more to the variance of allele frequency changes than drift, yet
across-taxa patterns in diversity are better explained processes acting
sporadically on longer timescales, such as colonization, founder effects, and
bottlenecks. Thus, the pairwise diversity may not give us the best picture of
the generation to generation evolutionary processes acting in a population to
change allele frequencies. Furthermore, certain observed adaptations occur at a
pace that is inexplicable given small effective population sizes implied by
diversity, and are only possible if short-term effective population sizes are
orders of magnitude larger \citep{Karasov2010-fb,Barton2010-pi}.

@@

In \emph{Building a Science of Population Biology}
(\citeyear{Lewontin2004-fb}), Lewontin laments the difficulty of uniting
population genetics and population ecology into a cohesive discipline of
population biology. Lewontin's Paradox of Variation remains a major unsolved
problem at the nexus of these two different disciplines: we fail to understand
the processes that connect a central parameter of population ecology, census
size, to a central parameter of population genetics, effective population size
across species. Given that selection seems to fall short in resolving
Lewontin's Paradox, a full resolution will require a mechanistic understanding
the ecological, life history, and macroevolutionary processes that connect
$N_c$ to $N_e$ across taxa. While I have focused exclusively on metazoan taxa
since their population densities are more readily approximated from body mass,
a full resolution must also include plant species (with the added difficulties
of variation in selfing rates, different dispersal strategies, pollination,
etc.).

Looking at Lewontin's Paradox through an macroecological and macroevolutionary
lens begets interesting questions outside of the traditional realm of
population genetics.  Here, I have found that diversity and $N_c$ have a
consistent relationship without many outliers, despite the wildly disparate
ecologies, life histories, and evolutionary histories of the taxa included.
Furthermore, taxa with very large census sizes have surprisingly low diversity.
Is this explained by macroevolutionary processes, such as different rates of
speciation for large-$N_c$ taxa? Or, are the levels of diversity we observe
today an artifact of our timing relative to the last glacial maximum, or the
last major extinction? Did large-$N_c$ prehistoric animal populations living in
other geological eras have higher levels of diversity than our present taxa?
Or, does ecological competition occur on shorter timescales such that strong
population size contractions transpire and depress diversity, even if a species
is undisturbed by climatic shifts or mass extinctions? Overall, patterns of
diversity across taxa are determined by many overlaid evolutionary and
ecological processes occurring on vastly different timescales. Lewontin's
Paradox of Variation may persist unresolved for some time because the
explanation requires synthesis and model building at the intersection of all
these disciplines.

@@

\begin{align}
  \E(\pi) &= \frac{\theta}{\theta + \nicefrac{1}{B} + 2NS} \\
          &\approx \frac{\theta}{\nicefrac{1}{B} + 2NS} 
\end{align}
%
(cf. equation 1 \cite{Elyashiv2016-vt}, equation 4 of \cite{Kim2000-ep}, and equation 20 of \cite{Coop2012-cd}, ).
The BGS component is given by \cite{Hudson1995-xc},

@@

\begin{align}
  S = \frac{\nu_\text{BP}}{r_\text{BP}} J
\end{align}
%
(cf. \cite{Coop2012-cd} equation 20) where $\nu_\text{BP}$ and $r_\text{BP}$
are the substitutions and recombination per basepair respectively, $J$ is the
probability that two lineages coalesce down to one, given sweeps occur
uniformly along the genome.  Under this homogeneous sweep model, $J$ is

@@

\begin{align}
  S(x) = \frac{1}{T} \sum_{i_S} \alpha(i_S) \sum_{y \in a(i_S)} \int \exp(-r(x, y) \tau(s, N)) g(s | i_S) ds
\end{align}
%
where $T$ is the length of the lineage (in generations) on which substitutions
accrue, $i_S = 1, \ldots, I_S$ is the annotation class (e.g. exons, introns,
UTRs), $\alpha(i_S)$ is the fraction of substitutions in annotation class $i_S$
that are beneficial, $a(i_S)$ is the set of all substitutions in annotation
class $i_S$, $\tau(s, N)$ is the fixation time of a site with additive effect
$s$, and $g(s | i_S)$ is the distribution of selection coefficients for
annotation class $i_S$.
